Analyzing Housing Affordability in the United States: The Role of Inflation, Interest Rates, and Supply

### 2024-11-25 paper/paper.qmd
finished intro, abstract, title, subtitle

---
title: "My title"
subtitle: "My subtitle if needed"
author: 
  - Krishna Kumar
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Overview paragraph

Estimand paragraph

Results paragraph

Why it matters paragraph

Telegraphing paragraph: The remainder of this paper is structured as follows. @sec-data....

@@

# Model

@@

# Results

>>>>

---
title: "Analyzing Housing Affordability in the United States: The Role of Inflation, Interest Rates, and Supply"
subtitle: "Rising Inflation and Interest Rates Worsen Affordability, While Increasing Housing Supply Offers Relief"
author: 
  - Krishna Kumar
thanks: "Code and data are available at: [https://github.com/RohanAlexander/starter_folder](https://github.com/RohanAlexander/starter_folder)."
date: today
date-format: long
abstract: "This paper investigates housing affordability in the United States from 2006 to 2024 by constructing an Affordability Index that measures the number of work hours required to purchase a median-priced home outright. Using a multiple linear regression model, we analyze the relationships between housing affordability and three key macroeconomic variables: Consumer Price Index (CPI), housing supply, and interest rates. Our findings reveal that rising inflation and interest rates significantly reduce affordability, while increased housing supply mitigates this effect. These insights highlight the complex interplay of economic forces shaping the housing market and emphasize the importance of stabilizing inflation, increasing supply, and maintaining balanced monetary policy to improve housing accessibility for U.S. households."
format: pdf
number-sections: true
bibliography: references.bib
---

@@

Housing affordability is a growing issue in the United States, with significant implications for economic stability and social well-being of individuals. Over the past two decades, escalating home prices, inflation, and shifting interest rates have exacerbated housing affordability issues, leaving many households unable to achieve home ownership. As policymakers seek solutions to these challenges, understanding the factors that drive housing affordability has become increasingly critical. This paper addresses this issue by focusing on the interplay of key macroeconomic variables—Consumer Price Index (CPI), housing supply, and interest rates—and their collective influence on housing affordability over time.

To capture the complexity of this issue, we develop an Affordability Index, a measure representing the number of work hours required to purchase a median-priced home outright. This index provides a clear, intuitive way to assess affordability trends int he long run and evaluate the impact of inflation, housing supply, and monetary policy on home ownership accessibility. Using multiple linear regression, we analyze how these factors interact and contribute to changes in affordability from 2006 to 2024. By systematically building and comparing three models, we examine the power of each predictor (marginally and collectively) and evaluate their significance in shaping affordability dynamics.

The findings of this paper reveal that inflation and interest rates significantly reduce housing affordability (i.e. increase the affordability index), while an increase in housing supply improves it. Rising CPI drives up the cost of living and housing prices, decreasing purchasing power. Similarly, higher interest rates inflate mortgage costs, further straining affordability. Conversely, increased housing supply alleviates upward pressure on prices, demonstrating its critical role in improving affordability. These results offer valuable insights for policymakers, highlighting the need for measures that stabilize inflation, increase housing stock, and balance monetary policy to enhance housing accessibility.

The remainder of this paper is organized as follows. Section @sec-data presents the data sources, variable descriptions, and exploratory analysis. Section @sec-model outlines the model details, setup, and justification. Section @sec-results highlights the key findings from the models and visualizations. Finally, Section @sec-discussion delves into the broader implications of these findings, addresses limitations, and suggests directions for future research. Lastly Section **(FIGURE OUT HOW TO REFERENCE APPENDIX HERE)** presents the model diagnostics and a survey, sampling, and observation method that would be ideal for data of the kind we used.

## Estimand

The estimand in this paper is the Affordability Index, representing the hours of work required to purchase a median-priced home outright. This measure allows us to quantify and analyze the impact of macroeconomic variables on affordability in a clear and meaningful way, especially when the analysis focuses on the long run. By focusing on this estimand, we contribute to a deeper understanding of housing market dynamics and provide actionable insights for addressing affordability challenges.

@@

# Model {#sec-model}

@@

# Results {#sec-results}
